Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-11 article/article.tex
picture

\section{Methods}
\label{sec:methods}
We now continue by discussing the methods applied in our research. The general approach taken for each of the algorithms discussed consists of the following steps:
\begin{enumerate}
  \item Preprocessing
  \item Dimensionality reduction
  \item Training
  \item Testing
\end{enumerate}
Preprocessing was done to transform the dataset into a format suitable for input into the classification algorithm. Dimensionality reduction was required to cope with the high dimensionality of the dataset and may consist of both feature extraction as well as feature selection. Finally, the approach taken to training and testing the classifier deals with preventing overfitting the classifier to training data and verifying accuracy for cross-subject classification. The choices made in each of these steps may affect classifier accuracy, so to obain a complete understanding of classification accuracies we have tried several methods at each step. In the following sections, these methods will be described in more detail.\\
The implementation of the algorithms was done using Python combined with modules tailored to mathematics (NumPy, SciPy) and machine learning (ScikitLearn). The original dataset was made available as a MatLab file and we used preprocessing in MatLab to create a number of files containing modified versions of this data, such as normalization and ROI grouping (see section~\ref{ssec:preprocessing}). The source code for the algorithms is available from the authors on request.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As with preprocessing, other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability and were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (see Fig. 1). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans prWindows API error 87: The parameter is incorrect.ovide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As can be seen in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.      
\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average BOLD signal for all subjects is plotted against the scan number. The coloured regions depict the mean BOLD signal +/- 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

Table~\ref{tab:results} illustrates the overall performance of the three classifiers. Rows indicate various feature selection methods. We applied PCA to the original data, while the rest feature selection methods were applied to the data belonging to the selected ROIs (cf.~Section~\ref{sec:dimensionalityReduction}). For instance, \emph{Norm/Vox-M ROIs} indicates that we normalised over the voxel values of the same voxel index using the mean and standard deviation, while \emph{Norm-M-SD ROIs} indicates that we normalised over the voxel values of the same scan index using the mean and standard deviation. Columns depict the used classifiers. For the classifiers depicted with *, we used a different method for structuring the data, as described in Section~\ref{sec:trainingMethodsAndClassifiers}.

Furthermore, to test the significance of the differences in performance between the different methods we performed a Chi-square test for interesting contrasts, and obtained a confidence interval for their difference in proportions. For the rest of this section we focus only on the case when \emph{Norm/Vox-M ROIs} is used as a feature selection method.

From Table~\ref{tab:results} we can distinguish the fact that when PCA is used for feature extraction, the performance of GNB is lower than when other feature selection methods are used. By using an exact binomial test we found that GNB that used PCA as feature extraction method performed significantly better than chance-level, with a 95\% $CI: 56.9 - 65.8$, $p < 0.001$. However, a Chi-square test for equality of proportions indicated that using PCA as dimentionality reduction method performed significantly worse than a GNB classifier on the means of the selected ROI's, $\chi^2(1) = 7.7848, p < 0.01$. This suggests that maximizing the variance of the data does not maximize information, possibly because the low signal-to-noise ratio of fMRI data does not insure that the principle component with the highest variance also contains a lot of useful information. Additionally, the data may not suited to be aligned by voxel index (see section~\ref{sec:dimensionalityReduction}). The performance listed in Table~\ref{tab:results} is for PCA using 149 components. This value was found by performing classification with a varying amount of components, the results for which are listed in Figure~\ref{fig:pca_components}.

@@

Comparing classifiers when the same data structure is used, we can distinguish the fact that MVG performs better than GNB. However, a Chi-square test did not find this difference to be significant, $\chi^2(1) = 0.8807, p = 0.348$. Furthermore, the difference between LSVM and GNB was also not found to be significant, $\chi^2(1) = 1.4183, p = 0.234$.

Finally, when classifying per scan and combining these classifications into a single classification per trial (cf.~Table~\ref{tab:results} indicated with a *) the performance increases. When looking at the increase in performance for the GNB classifier, we see performance increases from 73.13\% to 76.88\%. However, a Chi-square test did not find this increase in performance to be significant, $\chi^2(1) = 1.6056, p = 0.205$.

@@

In this paper we have replicated earlier findings of decoding the cognitive states of looking at sentences or pictures from fMRI data. Specifically, by observing the patterns of activity for a group of subjects, we are able to predict with above-chance accuracy what the cognitive state of an unseen subject is. This shows that machine learning algorithms can be employed to capture useful information from noisy data, and generalise this information between subjects.\\
\indent We found that even though all of our classification procedures yielded well above chance-level classification performance, the difference between the classifiers was seldomly significant. Still, we consistently found that the Multivariate Gaussian (MVG) classifier outperformed the Gaussian Naive Bayes (GNB) classifier and Linear Support Vector Machine (LSVM). This finding suggests that the interactions between different brain areas contain information about the cognitive state, and that these functional connectivity generalises to an extent to different subjects.\\
\indent Still, the uncertainty in our predictions are relatively high, with maximum classification performance reaching close to 80\% accuracy for only two classes. The noisy nature of the fMRI data therefore makes it very difficult to draw hard conclusions when decoding a single cognitive state from a brain scan. This is something to be considered before employing fMRI as a diagnostic tool, or as evidence in court.\\
\indent Overall we can say that machine learning provides a feasible way of extracting useful information from fMRI data. Furthermore, it avoids a strict multiple comparisons correction that is required in null hypothesis testing by using cross-validation. Also, the identification of patterns of activity associated with tasks is encouraged, rather than attempting to identifying brain regions that are involved in tasks. In doing this, much can be learned about how information is represented and processed in the brain.

>>>>

\section{Methods}
\label{sec:methods}
We now continue by discussing the methods applied in our research. The general approach taken for each of the algorithms discussed consists of the following steps:
\begin{enumerate}
  \item Preprocessing
  \item Dimensionality reduction
  \item Training
  \item Testing
\end{enumerate}
Preprocessing was done to transform the dataset into a format suitable for input into the classification algorithm. Dimensionality reduction was required to cope with the high dimensionality of the dataset and may consist of both feature extraction as well as feature selection. Finally, the approach taken to training and testing the classifier deals with preventing overfitting the classifier to training data and verifying accuracy for cross-subject classification. The choices made in each of these steps may affect classifier accuracy, so to obain a complete understanding of classification accuracies we have tried several methods at each step. In the following sections, these methods will be described in more detail.\\
The implementation of the algorithms was done using Python combined with modules tailored to mathematics (NumPy, SciPy) and machine learning (ScikitLearn). The original dataset was made available as a MatLab file and we used preprocessing in MatLab to create a number of files containing modified versions of this data, such as normalization and ROI grouping (see section~\ref{preprocessing}). The source code for the algorithms is available from the authors on request.

@@

\textbf{\emph{ROI mapping.}} ROI mapping is a form of feature selection, wherein only a subset of the N-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. We used 7 of the 25 defined ROIs, labeled  'CALC' 'LIPL' 'LT' 'LTRIA' 'LOPER' 'LIPS' and  'LDLPFC'. These ROIs have previously been found to be most relevant in discriminating between the cognitive states of looking at sentences or pictures \cite{wa:2003betw}. To further reduce the number of features, average voxel values within these regions are taken as new input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis. The disadvantage of using ROIs is that they require \textit{a priori} knowledge. When this knowledge is inaccurate, classification accuracy may not be optimal.

@@

\textbf{\emph{Temporal dimensionality reduction.}} The BOLD response peaks at approximately 5 seconds after a stimulus is presented (see Fig. 1). This means that the first scans after a stimulus may not represent activity related to this stimulus and should not be used for classification. To examine which scans provide maximal class information, we plotted the average BOLD activity per scan for all trials of all subjects, given a trial type and ROI. As can be seen in Figure~\ref{fig:boldroi}, scans 10 to 20 provide maximal information for the first stimulus, while scans 23 to 33 provide maximal information for the second stimulus. The other scans were excluded.      
\begin{figure}
	\centering
	\includegraphics[width=90mm]{figures/scan_hemodynamic_BOLD_roi}
  	\caption{Haemodynamic response for ROI "CALC". The red line depicts a sentence-picture trial (S-P) and the blue line depicts a picture-sentence trial (P-S). Average BOLD signal for all subjects is plotted against the scan number. The coloured regions depict the mean BOLD signal +/- 1 standard deviation.}
  	\label{fig:boldroi}
\end{figure}

@@

Table~\ref{tab:results} illustrates the overall performance of the three classifiers. Rows indicate various feature selection methods. We applied PCA to the original data, while the rest feature selection methods were applied to the data belonging to the selected ROIs (cf.~Section~\ref{sec:dimensionalityReduction}). For instance, \emph{Norm/Vox-M-SD ROIs} indicates that we normalised over the voxel values of the same voxel index using the mean and standard deviation, while \emph{Norm-M-SD ROIs} indicates that we normalised over the voxel values of the same scan index using the mean and standard deviation. Columns depict the used classifiers. For the classifiers depicted with *, we used a different method for structuring the data, as described in Section~\ref{sec:trainingMethodsAndClassifiers}.

Furthermore, to test the significance of the differences in performance between the different methods we performed an exact binomial test in each case, from which we obtained the confidence interval. The results were rounded to one decimal point. For the rest of this section and to explain some of the results we are looking only at the case when \emph{Norm/Vox-M-SD ROIs} is used as a feature selection method.

From Table~\ref{tab:results} we can distinguish the fact that when PCA is used for feature extraction, the performance of GNB is lower than when other feature selection methods are used. The confidence interval of the predictive performance of GNB when PCA is used is 95\% $CI: 56.9 - 65.8$ which is significantly worse than applying GNB in combination with other feature selection methods. This suggests that maximizing the variance of the data does not maximize information. Another explanation could be that the data is not suited to be aligned by voxel index (see section~\ref{sec:dimensionalityReduction}). The performance listed in Table~\ref{tab:results} is for PCA using 149 components. This value was found by performing classification with a varying amount of components, the results for which are listed in figure~\ref{fig:pca_components}.

@@

Comparing classifiers when the same data structure is used, we can distinguish the fact that MVG performs better than GNB. Also that LSVM performs better than GNB, though not significantly. The confidence interval of GNB is 95\% $CI: 68.9-77.0$ and of LSVM is 95\% $CI: 72.6-80.4$, which suggests that their performances do not differ significantly.

Moreover, when different method for structuring the data is used the performance increases, though not significantly. This can be seen by comparing the performance of GNB when different method for structuring the data is used. In the first case it is 73.13\% while in the second case is 76.88\%. The confidence interval in the first case is 95\% $CI: 68.9 - 77.0$ and in the second case is 95\% $CI: 72.8 - 80.6$.

Finally, we can distinguish the fact that MVG outperforms all the other classifiers, though we did not try a different data structure in the case of LSVM. However, we still do not expect that the performance of LSVM will be higher that this of MVG. The confidence interval of MVG is 95\% $CI: 75.7-83.1$.

@@

In this paper we have replicated earlier findings of decoding the cognitive states of looking at sentences or pictures from fMRI data. Specifically, by observing the patterns of activity for a group of subjects, we are able to predict with above-chance accuracy what the cognitive state of an unseen subject is. This shows that machine learning algorithms can be employed to capture useful information from noisy data, and generalise this information between subjects.\\
\indent We found that even though all of our classification procedures yielded well above chance-level classification performance, the differences between the classifiers was very seldom significant. Still, we consistently found that the Multivariate Gaussian (MVG) classifier outperformed the Gaussian Naive Bayes (GNB) classifier and Linear Support Vector machine (LSV). This finding suggests that the interactions between different brain areas contain information about the cognitive state, and that these functional connectivity generalises to an extent to different subjects.\\
\indent Still, the uncertainty in our predictions are relatively high, with maximum classification performance reaching close to 80\% accuracy for only two classes. The noisy nature of the fMRI data therefore makes it very difficult to draw hard conclusions when decoding a single cognitive state from a brain scan. This is something to be considered before employing fMRI as a diagnostic tool, or as evidence in court.\\
\indent Overall we can say that machine learning provides a feasible way of extracting useful information from fMRI data. Furthermore, it avoids a strict multiple comparisons correction that is required in null hypothesis testing by using cross-validation. Also, the identification of patterns of activity associated with tasks is encouraged, rather than attempting to identifying brain regions that are involved in tasks. In doing this, much can be learned about how information is represented and processed in the brain.
